---
title: "Bivariate Analysis in R"
output:
  word_document: default
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
---
Bivariate Analysis is an integral part of the EDA process. It helps us explore the relationship between different variables and discover valuable insights from the data.
Here we show visualization methods to perform bivariate analysis based on the type of variables.
There are different methods for visualizing Quantitative (continuous) and Categorical(discrete) variables.

All the examples are demonstrated using the mpg dataset in ggplot2

This document is based on a [book](https://rkabacoff.github.io/datavis) by Rob Kabacoff. Please feel free to visit his website for more data visualization techniques in R.


```{r include = FALSE}
#installing the required packages
# install.packages("ggplot2")
# install.packages("scales")
# install.packages("ggbeeswarm")
```
*** 
# Quantitative vs Quantitative
Scatter plots are a good way to showcase the relation between two quantitative variables

## Scatter Plot
```{r echo=TRUE} 
library(ggplot2) #loading the library required for this plot
#scatter plot
ggplot(mpg,aes(x = displ,y = hwy)) + 
  geom_point(color="cornflowerblue", #setting the colour, size and transparency(alpha) of the points
             size = 2, 
             alpha=.8) +
  scale_x_continuous(breaks = seq(1, 7, 1), #customizing x-axis : seq(start of units, end of units, size of one unit)
                     limits = c(1, 7)) + #setting limits of the axis
  scale_y_continuous(breaks = seq(10, 45, 5), 
                     limits=c(10, 50)) +
  labs(x = "Displacement in Litres", #specifying the labels of axes and title of plot
       y = "Highway fuel economy (miles per gallon)",
       title = "mpg vs Displacement",
       subtitle = "Relation between engine size and mpg") +
  geom_smooth(method = "lm") # this adds a linear trend line which is useful to summarize the relationship between the two variables
```
<br />
Here we can see that fuel economy has an inverse relationship with displacement

***
# Categorical vs Quantitative
Kernal Density Plots, Box plots, Jitter plots and Beeswarm plots can be used to explain relationship between Categorical and Quantitative variables in a beautiful and effective manner

## Kernal Density Plot
```{r echo= TRUE}
#kernel density plots
ggplot(mpg, 
       aes(x = cty, #quantitative variable
           fill = factor(drv, #defining x axis a categorical
                         levels = c("f", "r", "4"),
                         labels = c("front-wheel", #labelling the legend
                                    "rear-wheel", 
                                    "4-wheel")))) +
  geom_density(alpha = 0.4) + #setting transparency of graph to keep overlaps visible
  labs(fill = "Drive Train", # setting title of legend
       x = "mpg",
       title = "Fuel Economy by Drive type")
```
<br />
Here we can see that front-wheel drive vehicles generally have higher fuel economy and are also more spread than the other two drive types

## Jitter Box Plot
```{r echo =TRUE}
#jitter and box plots
library(scales)
ggplot(mpg, 
       aes(x = factor(drv, #defining x axis a categorical
                      labels = c("rear-wheel", #labelling the axis with categorical variable
                                    "front-wheel", 
                                    "4-wheel")), 
           y = cty, 
           color = drv)) + #specifying that coloring is to be based on drive type
  geom_boxplot(size=1, #makes the lines thicker
               outlier.shape = 1, #specifies circles for outliers
               outlier.color = "black", #makes outliers black
               outlier.size  = 3) + #increases the size of the outlier symbol
  geom_jitter(alpha = 0.5, #setting transparency of graph
              width=.2) + #decreases the amount of jitter (.4 is the default)
  labs(title = "Fuel Economy by Drive type", 
       x = "",
       y = "mpg") +
  theme_minimal() + #setting minimal theme (no background color)
  theme(legend.position = "none") + #hiding legend
  coord_flip() #x and y axes are reversed
```
<br /> 


## Beeswarm Plot
```{r warning=FALSE}
#beeswarm  plot
library(ggbeeswarm)
library(scales)
ggplot(mpg, 
       aes(x = factor(drv,
                      labels = c("rear-wheel", #labelling the axis with categorical variable
                                    "front-wheel", 
                                    "4-wheel")), 
           y = cty, 
           color = drv)) +
  geom_quasirandom(alpha = 0.7, 
                   size = 1.5) + #setting size of dots
  labs(title = "Fuel Economy by Drive type", 
       x = "drive type",
       y = "mpg") +
  theme_minimal() +
  theme(legend.position = "none")
```
<br />
 
***
# Categorical vs Categorical
Bar charts provide a simple and elegant way to represent categorical vs categorical data visually

## Stacked bar chart
```{r echo=TRUE}
# stacked bar chart
ggplot(mpg, 
       aes(x = class, 
           fill = factor(drv, 
                         levels = c("f", "r", "4"),
                         labels = c("front-wheel", #labelling the legend
                                    "rear-wheel", 
                                    "4-wheel")))) + 
  labs(fill = "Drive Train", # setting title of legend
       x = "vehicle class",
       title = "Automobile drive type by class") +
  geom_bar(position = "stack") #specifying the type of bar chart as stacked
```
  
## Grouped Bar Chart
```{r echo=TRUE}
#Grouped bar chart
ggplot(mpg, 
       aes(x = class, 
           fill = factor(drv, 
                         levels = c("f", "r", "4"),
                         labels = c("front-wheel", #labelling the legend
                                    "rear-wheel", 
                                    "4-wheel")))) + 
  labs(fill = "Drive Train", # setting title of legend
       x = "vehicle class",
       title = "Automobile drive type by class") +
geom_bar(position = position_dodge(preserve = "single")) #specifying that grouped bar chart should have 3 lines for every class
```
  
## Segmented Bar Chart
```{r echo=TRUE}
#Segmented bar chart
ggplot(mpg, 
       aes(x = class, 
           fill = factor(drv, 
                         levels = c("f", "r", "4"),
                         labels = c("front-wheel", #labelling the legend
                                    "rear-wheel", 
                                    "4-wheel")))) + 
  labs(fill = "Drive Train", # setting title of legend
       x = "vehicle class",
       title = "Automobile drive type by class") +
geom_bar(position = "fill") + #specifying that the bar chart is of segmented type
  labs(y = "Proportion")
```


